Component response rate variation underlies stability in large complex systems

### 2019-03-01 notebook/ms.Rmd
Major restructure and revision of the Discussion to compare to Gibbs et al.

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, realistic complex systems (such that $S$ is high but finite) in which $\gamma$ varies are actually more likely to be stable.  <!--- Need to emphasise that this is for *real* systems at the thresholds of stability --->

Introduction could end here.

@@

It is important to emphasise that variation in component response rate is not stabilising per se; that is, adding variation in component response rate to a particular system does not necessarily increase the probability that the system will be stable. Rather, systems that are observed to be stable are more likely to have varying component response rates, and for this variation to be critical to their stability (Fig. 3). This is caused by the shift to a non-uniform distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes cause all of the real components of the eigenvalues of the community matrix $\mathbf{M}$ to become negative, but might also increase the real components of eigenvalues. The mathematics underlying this shift in eigenvalue distribution has been investigated [@Ahmadian2015] and recently applied to questions concerning species density and feasibility [@Gibbs2017; @Stone2017], but has not been interpreted as rates of response of individual system components to perturbation.

Also important to emphasise Gibbs result -- I'm doing this for finite systems, and I deliberate stressed the system complexity to the point of collapse.

But Gibbs was more interested in first assuming a stable matrix and then showing that the vector of abundances would not change it as S \to infinity -- Instead of assuming the system was stable prior to adding component response rate, I allow both the system and component response rate to vary to see the effects on stability.

@@

My results show that complex systems are more likely to be stable when the response rates of system components vary. These results are broadly applicable to understanding stability of complex networks in the physical, life, and social sciences.

>>>>

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, realistic complex systems (such that $S$ is high but finite) in which $\gamma$ varies are actually more likely to be stable.

@@

It is important to emphasise that variation in component response rate is not stabilising per se; that is, adding variation in component response rate to a particular system does not necessarily increase the probability that the system will be stable. Rather, highly complex systems that are observed to be stable are more likely to have varying component response rates, and for this variation to be critical to their stability (Fig. 3). This is caused by the shift to a non-uniform distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes cause all of the real components of the eigenvalues of the community matrix $\mathbf{M}$ to become negative, but might also increase the real components of eigenvalues. In this way, my focus is distinct from that of Gibbs et al. [@Gibbs2017], who applied the same mathematical framework to investigate how a diagonal matrix $\mathbf{X}$ (equivalent to $\gamma$ in my model) affects the stability of a community matrix $\mathbf{M}$ given an interaction matrix $\mathbf{A}$ within a generalised Lotka-Volterra model, where $\mathbf{M} = \mathbf{XA}$. Gibbs et al. [@Gibbs2017] analytically demonstrated that the effect of $\mathbf{X}$ on system stability decreases exponentially as system size becomes arbitrarily large ($S \to \infty$). My numerical results are entirely consistent with this prediction (see Supplementary Information). Given that a complex system is known to be stable (or unstable), $Var(\gamma)$ is not expected to increase or decrease the probability of stability. Rather, component response rate variation increases the upper bound at which stability can be realistically observed, meaning that highly complex systems are more likely than not to vary in their component response rates, and for this variation to be critical for system stability.

@@

A general mathematical framework encompassing shifts in eigenvalue distributions caused by a vector $\gamma$ has been investigated [@Ahmadian2015] and recently applied to questions concerning species density and feasibility [@Gibbs2017; @Stone2017], but $\gamma$ has not been interpreted as rates of response of individual system components to perturbation. My model focuses on component response rates for systems of a finite size, in which complexity is high but not yet high enough to make the probability of stability unrealistically low for actual empirical systems. For this upper range of system size, randomly assembled complex systems are more likely to be stable if their component response rates vary (e.g., $10 < S < 30$ for parameter values in Fig. 3). Overall, I suggest that variation in component response rate might therefore be critical for maintaining stability in many highly complex empirical systems. These results are broadly applicable to understanding stability of complex networks across the physical, life, and social sciences.

### 2019-03-01 notebook/ms.tex
Major restructure and revision of the Discussion to compare to Gibbs et al.

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma\sqrt{SC} < 1\) assumes that the expected
response rates (\(E[\gamma]\)) of individual components to perturbations
of the system are identical, but this is highly unlikely in any complex
system. In ecological communities, for example, the rate at which
population density changes following perturbation will depend on the
generation time of organisms, which might vary by orders of magnitude
among species. Species with short generation times will respond quickly
(high \(\gamma\)) to perturbations relative to species with long
generation times (low \(\gamma\)). Similarly, the speed at which
individual banks respond to perturbations in financial networks, or
individuals or institutions respond to perturbations in complex social
networks, is likely to vary. The effect of such variance on stability
has not been investigated in complex systems theory. Intuitively,
variation in \(\gamma\) might be expected to decrease system stability
by introducing a new source of variation into the system and thereby
increasing \(\sigma\). Here I show why, despite higher \(\sigma\),
realistic complex systems (such that \(S\) is high but finite) in which
\(\gamma\) varies are actually more likely to be stable.

@@

It is important to emphasise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rate to a particular system does not necessarily increase the
probability that the system will be stable. Rather, systems that are
observed to be stable are more likely to have varying component response
rates, and for this variation to be critical to their stability (Fig.
3). This is caused by the shift to a non-uniform distribution of
eigenvalues that occurs by introducing \(Var(\gamma)\) (Fig. 1b, 2b),
which can sometimes cause all of the real components of the eigenvalues
of the community matrix \(\mathbf{M}\) to become negative, but might
also increase the real components of eigenvalues. The mathematics
underlying this shift in eigenvalue distribution has been
investigated\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{16}}
and recently applied to questions concerning species density and
feasibility\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17},\protect\hyperlink{ref-Stone2017}{18}},
but has not been interpreted as rates of response of individual system
components to perturbation.

@@

My results show that complex systems are more likely to be stable when
the response rates of system components vary. These results are broadly
applicable to understanding stability of complex networks in the
physical, life, and social sciences.

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.719 and 0.737 for a and b,
respectively.

>>>>

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma\sqrt{SC} < 1\) assumes that the expected
response rates (\(\gamma\)) of individual components to perturbations of
the system are identical, but this is highly unlikely in any complex
system. In ecological communities, for example, the rate at which
population density changes following perturbation will depend on the
generation time of organisms, which might vary by orders of magnitude
among species. Species with short generation times will respond quickly
(high \(\gamma\)) to perturbations relative to species with long
generation times (low \(\gamma\)). Similarly, the speed at which
individual banks respond to perturbations in financial networks, or
individuals or institutions respond to perturbations in complex social
networks, is likely to vary. The effect of such variance on stability
has not been investigated in complex systems theory. Intuitively,
variation in \(\gamma\) might be expected to decrease system stability
by introducing a new source of variation into the system and thereby
increasing \(\sigma\). Here I show why, despite higher \(\sigma\),
realistic complex systems (such that \(S\) is high but finite) in which
\(\gamma\) varies are actually more likely to be stable.

@@

It is important to emphasise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rate to a particular system does not necessarily increase the
probability that the system will be stable. Rather, highly complex
systems that are observed to be stable are more likely to have varying
component response rates, and for this variation to be critical to their
stability (Fig. 3). This is caused by the shift to a non-uniform
distribution of eigenvalues that occurs by introducing \(Var(\gamma)\)
(Fig. 1b, 2b), which can sometimes cause all of the real components of
the eigenvalues of the community matrix \(\mathbf{M}\) to become
negative, but might also increase the real components of eigenvalues. In
this way, my focus is distinct from that of Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\gamma\) in my model) affects the
stability of a community matrix \(\mathbf{M}\) given an interaction
matrix \(\mathbf{A}\) within a generalised Lotka-Volterra model, where
\(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)). My numerical results are entirely consistent with
this prediction (see Supplementary Information). Given that a complex
system is known to be stable (or unstable), \(Var(\gamma)\) is not
expected to increase or decrease the probability of stability. Rather,
component response rate variation increases the upper bound at which
stability can be realistically observed, meaning that highly complex
systems are more likely than not to vary in their component response
rates, and for this variation to be critical for system stability.

@@

A general mathematical framework encompassing shifts in eigenvalue
distributions caused by a vector \(\gamma\) has been
investigated\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{16}}
and recently applied to questions concerning species density and
feasibility\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17},\protect\hyperlink{ref-Stone2017}{18}},
but \(\gamma\) has not been interpreted as rates of response of
individual system components to perturbation. My model focuses on
component response rates for systems of a finite size, in which
complexity is high but not yet high enough to make the probability of
stability unrealistically low for actual empirical systems. For this
upper range of system size, randomly assembled complex systems are more
likely to be stable if their component response rates vary (e.g.,
\(10 < S < 30\) for parameter values in Fig. 3). Overall, I suggest that
variation in component response rate might therefore be critical for
maintaining stability in many highly complex empirical systems. These
results are broadly applicable to understanding stability of complex
networks across the physical, life, and social sciences.

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.724 and 0.738 for a and b,
respectively.
